---
output:
  md_document:
    variant: markdown_github
  html_document: default
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# sparseEigen

This package provides two functions to compute sparse eigenvectors (while keeping their orthogonality property) from either the covariance matrix or directly the data matrix.


## Installation
```{r, eval = FALSE}
# Installation from local file
install.packages(file.choose(), repos = NULL, type="source")

# Or from GitHub
# install.packages("devtools")
devtools::install_github("dppalomar/sparseEigen")

# Get help
library(sparseEigen)
help(package="sparseEigen")
?spEigen
```


## Usage

The following is a simple illustrative example.

We start by loading the package and generating synthetic data with sparse eigenvectors:
```{r}
library(sparseEigen)

# parameters 
m <- 500  # dimension
n <- 100  # number of samples
q <- 3  # number of sparse eigenvectors to be estimated
sp_card <- 0.2*m  # cardinality of each sparse eigenvector
rho <- 0.6  # sparsity level

# generate non-overlapping sparse eigenvectors
V <- matrix(rnorm(m^2), ncol = m)
tmp <- matrix(0, m, q)
for (i in 1:max(q, 2)) {
  ind1 <- (i - 1)*sp_card + 1
  ind2 <- i*sp_card
  tmp[ind1:ind2, i] = 1/sqrt(sp_card)
  V[, i] <- tmp[, i]
}
V <- qr.Q(qr(V))  # keep first q eigenvectors the same (already orthogonal) and orthogonalize the rest

# generate eigenvalues
lmd <- rep(1, m)
lmd[1:q] <- 100*seq(from = q, to = 1)

# generate covariance matrix from sparse eigenvectors and eigenvalues
R <- V %*% diag(lmd) %*% t(V)

# generate data matrix from a zero-mean multivariate Gaussian distribution with the constructed covariance
X <- MASS::mvrnorm(n, rep(0, m), R)  # random data with underlying sparse structure
X <- scale(X, center = TRUE, scale = FALSE)  # center the data
```
Then we estimate the covariance matrix with `cov(X)` and compute its sparse eigenvectors:
```{r}
# computation of sparse eigenvectors
res_standard <- eigen(cov(X))
res_sparse <- spEigen(X, q, rho)
# res_sparse <- spEigenCov(cov(X), q, rho)
```

We can assess how good the estimated eigenvectors are by computing the inner product with the original eigenvectors (the closer to 1 the better):
```{r}
# show inner product between estimated eigenvectors and originals
abs(diag(t(res_standard$vectors) %*% V[, 1:q]))  #for standard estimated eigenvectors
abs(diag(t(res_sparse$vectors) %*% V[, 1:q]))    #for sparse estimated eigenvectors
```

Finally, the following plot shows the sparsity pattern of the eigenvectors:
```{r, echo = FALSE, fig.height = 7, fig.width = 7}
par(mfcol = c(3, 2))
plot(res_sparse$vectors[, 1]*sign(res_sparse$vectors[1, 1]), main = "First Sparse Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 1]*sign(V[1, 1]), col = "red")
plot(res_sparse$vectors[, 2]*sign(res_sparse$vectors[sp_card+1, 2]), main = "Second Sparse Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 2]*sign(V[sp_card+1, 2]), col = "red")
plot(res_sparse$vectors[, 3]*sign(res_sparse$vectors[2*sp_card+1, 3]), main = "Third Sparse Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 3]*sign(V[2*sp_card+1, 3]), col = "red")

plot(res_standard$vectors[, 1]*sign(res_standard$vectors[1, 1]), main = "First Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 1]*sign(V[1, 1]), col = "red")
plot(res_standard$vectors[, 2]*sign(res_standard$vectors[sp_card+1, 2]), main = "Second Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 2]*sign(V[sp_card+1, 2]), col = "red")
plot(res_standard$vectors[, 3]*sign(res_standard$vectors[2*sp_card+1, 3]), main = "Third Eigenvector", xlab = "Index", ylab = "", type = "h")
lines(V[, 3]*sign(V[2*sp_card+1, 3]), col = "red")
```



